Regularised parameter selection in the non-stationary Hawkes process

### 2017-04-28 _paper_output/chapter_sparse_hawkes.tex
notation updates

We are interested in the problem of simultaneous identification of the
parameters of a Hawkes point process and identifying an inhomogeneous
driving signal by means of \(\ell_1\) penalised regression. This is
analaogous to a blind deconvolution problem \cite{stockham_blind_1975},
adaptive robust regression \cite{donoho_automatic_1988}, or a sparse
inverse problem \cite{tibshirani_regression_1996}, or, but considerably
more challenging; While the process model superficially resembles a
linear system, there are few of the convenient results that would apply
in a linear regression problem; and as a time series problem it is
difficult to choose our penalty parameters by standard non-parametric
means such as cross validation. The proposed method uses a
heuristically-motivated degrees-of-freedom regularisation parameter
selection, which we test through simulation. The proposed method works
well in practice.

@@

We are given a counting process \(N\) on \(\mathbb R\), and an interval
\([0,T]\). (TODO clarify fit interval versus data interval)

We represent a realisation of the time series by its arrival times
\(\mathbf{t}=\{t_1, t_2,\dots,t_N(T)\}\). We may also, when as
notationally conveneient, use the equivalent representation as the
function that counts the arrivals:

\[
N(t):= \left|\{s:s\in \mathbf{t}\,\&\,s\leq t\}\right|
\]

We use only the left-continuous filtration
\(\mathcal{F}(\{N(s)\}_{s < t})\) generated by the process, which we
here write as \(\mathcal{F}(N)\)

We call it a (TODO) predictable Poisson point process if it has an
associated \emph{intensity} process \(\lambda\), such that \(\lambda\)
is predictable with respect to \(\mathcal{F}(N)\) generated by the
process, i.e.

\[\lambda(\mathcal{F}(t)):=\lim_{h\to 0} \frac{\mathrm E\left(N(t,t+h)\right)}{h}.\]
\cite{daley_introduction_2003}

    The Hawkes process \cite{hawkes_point_1971} in particular, is the
linear, self-exciting point process; Concretely, given a "background"
rate \(\mu:\mathbb{R}^+\rightarrow\mathbb{R}^+,\) and a non-negative
influence kernel with positive support
\(\phi:\mathbb{R}\rightarrow\mathbb{R}^+\cup\{0\}\), the Hawkes process
has the following expression for its intensity:

\[
\begin{aligned}
\lambda(\mathcal{F(N(t))}^-_{(-\infty,t)}) &= \mu(t) + \phi * N\\
&= \mu(t)  + \int_{-\infty}^{\infty}\phi(t-s)dN(s)\\
&= \mu(t) + \int_{-\infty}^{t}\phi(t-s)dN(s)\\
&= \mu(t) + \sum_{t_i\lt t}\phi(t-t_i)
\end{aligned}
\]

For brevity, we write
\(\lambda(\mathcal{F(t)}^-_{(-\infty,N(t))})=\lambda(t)\).

    The log-likelihood function for a given realisation of the data is given
\cite{ozaki_etimating_1979}

\[L_\pi(N;\mu,\phi):=-\int_0^T\lambda(t;\mu,\phi)dt + \int_0^T\log \lambda(t;\mu,\phi) dN_t \]

Using the method of maximum likelihood we estimate our parameter vector
by numerical optimisation of
\[\hat{\theta}_\pi(N) = \operatorname{argmax}_{\mu,\phi} L_\pi(N;\theta).\]

Similarly, we will approximate \(\phi\) in a different functional basis
\(\{\bar{\psi}\}_{1\leq k \leq K}\), such that
\(\phi(t):= \sum_{k=1}^K \bar{\psi}(t;k)\). This basis choice is likely
arbitrary, but many exist in the literature trading off computational
and statistical convenience, including Laguerre polynomials
\cite{ogata_linear_1982}, Exponential functions
\cite{schoenberg_consistent_2005,rambaldi_modeling_2015}, and step
functions \cite{eichler_graphical_2016} As a convenient choice, we
propose a basis of smooth unimodal non-negative functions with a
computationally convenient form, the Maxwell density,
\[\bar{\psi}_j(t;\sigma):=\sqrt{\frac{2}{\pi}} \frac{t^2\exp\left(-\frac{t^2}{2\sigma^2}\right)}{\sigma^3}.\]
so that

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{paper_sparse_hawkes_files/paper_sparse_hawkes_10_0.pdf}
    \end{center}
    { \hspace*{\fill} \\}

    Since this is a general nonparametric problem we will use traditional
regularisation techniques to control estimates. We hope to recover the
unknown excitation kernel and a background intensity by appropriate
penalisation of the terms.

For the penalized log likelihood for penalty functional \(\pi\), we
write

@@

@@

    \(\ell_1\) type penalties are known for their ability to produce sparse
solutions, so we choose penalties of the form
\[\pi(\theta)=p\left\|\{\sigma_j\}\right\|_1+q\left\|\{\omega_j\}\right\|_1.\]
With the right choice of the parameters \(p\) and \(q\) we hope to
shrink unnecessary parameters to zero, identifying predictively
necessary ones, as in Lasso selection.

    There is no closed-form solution for the optimum of the Hawkes model
likelihood, penalised or not. The derivatives of the unpenalised
likelihood are easy to calculate manually -\/-\/-
\cite{ozaki_maximum_1979}. Here, however, we use standard reverse-mode
automatic differentiation \cite{baydin_automatic_2014}. The optimisation
proceeds using a standard gradient-based black-box based Truncated
Newton algorithm; as more specialised sparse regression algorithms such
as Least-Angle Regression, do not directly apply here.

@@

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{paper_sparse_hawkes_files/paper_sparse_hawkes_23_0.pdf}
    \end{center}
    { \hspace*{\fill} \\}

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{paper_sparse_hawkes_files/paper_sparse_hawkes_23_1.pdf}
    \end{center}
    { \hspace*{\fill} \\}

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{paper_sparse_hawkes_files/paper_sparse_hawkes_23_2.pdf}
    \end{center}
    { \hspace*{\fill} \\}

@@

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{paper_sparse_hawkes_files/paper_sparse_hawkes_25_0.png}
    \end{center}
    { \hspace*{\fill} \\}

@@

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{paper_sparse_hawkes_files/paper_sparse_hawkes_27_0.png}
    \end{center}
    { \hspace*{\fill} \\}

>>>>

We are interested in the problem of simultaneous identification of the

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  self excitation kernel, and
\item
  (possibly inhomogenous) background rate
\end{enumerate}

of a Hawkes point process a using of \(\ell_1\) penalised regression.

This is analaogous to a blind deconvolution problem
\cite{stockham_blind_1975}, adaptive robust regression
\cite{donoho_automatic_1988}, or a sparse inverse problem
\cite{tibshirani_regression_1996}, or, but considerably more
challenging; While the process model superficially resembles a linear
system, there are few of the convenient results that would apply in a
linear regression problem; and as a time series problem it is difficult
to choose our penalty parameters by standard non-parametric means such
as cross validation. The proposed method uses a heuristically-motivated
degrees-of-freedom regularisation parameter selection, which we test
through simulation. The proposed method works well in practice.

@@

We are given a counting process \(N(t)\) on \(\mathbb R\)

We represent a realisation of the time series by a (finite by
assumption) sequence of arrival times
\(\mathbf{t}=\{\dots,t_1, t_2,\dots,t_N(T)\}.\) We will be concerned
with estimating its likelihood based on the likelihood of the series on
an interval an interval \([0,T]\). We define
\(\mathbf{t}':=\mathbf{t}\cap [0,T]\) and assume
\(\left|\mathbf{t}'\right|<\infty.\)

We also, when notationally convenient, use the equivalent representation
as the left-continuous function that increments at each arrival time: \[
N(t)-N(s):= \left|\{s:s\in \mathbf{t}\,\mathrm{and}\,s\leq t\}\right| \,\mathrm{ for }\, t\geq s
\]

We set \(\lim_{s\rightarrow 0} N(s)\equiv 0\) and
\(t_1\equiv \inf \mathbf{t}'\); Indices \(i\leq 0\) are reserved for
point in the "history" of the process before time \(0\), which may be
generated by any arbitrary process. We will not be concerned with
evaluating the likelihood of such histories but may retain them to
control for their influence on the future of the series. For the rates
to be well-defined, however, we will assume that the increment sof the
process are finite on any finite interval.

We use only the left-continuous filtration
\(\mathcal{F}(\{N(s)\}_{s < t})\) generated by the process, which we
here write as \(\mathcal{F}(N(t)).\)

We call it a regular point process if it has an associated
\emph{intensity} process \(\lambda\), such that \(\lambda\) is
predictable with respect to \(\mathcal{F}(N)\) generated by the process,
i.e. \cite{daley_introduction_2003}

\[\lambda(t|\mathcal{F}(N(t))):=\lim_{h\to 0} \frac{\mathrm E\left(N(t,t+h)\right)}{h}.\]

Hereafter we define \(\lambda^*(t):=\lambda(t|\mathcal{F}(N(t)))\) for
brevity.

Such regular point processes generalise the homogeneous Poisson rate in
the sense that if we choose \(\lambda*(t)\equiv \lambda_0\) then it is
the intensity of a homogeneous Poisson rate with fixed intensity
\(\lambda_0\).

    \[N(t)\sim \operatorname{Pois}(\Lambda(t|\mathcal{F}(N(t)))\] where
\(\Lambda(s,t|\mathcal{F}(N(t)))=\int_s^t \lambda(s|\mathcal{F}(N(s))).\)

The waiting times are exponentially distributed given the intensity,
i.e. their densities \(f_i(s)\) are given \[\begin{aligned}
f_i(t)&\equiv f(t_{i+1}-t_i|\mathcal{F}(N(t_i)))\\
&=\Lambda(t_i,t|\mathcal{F}(N(t_i))\exp\left(-t\Lambda(t_i,t|\mathcal{F}(N(t_i))\right)
\end{aligned}\]

We will use the for the inter-arrival waiting times \(s_i=t_{i+1}-t_i\)
to calculate the data likelihood, after Rubin \cite{rubin_regular_1972}.

    The waiting times are exponentially distributed given the intensity,
i.e. their densities \(f_i(s)\) for any given transition are given
\[\begin{aligned}
\frac{d}{dt}P(t_{i+1}=t|\mathcal{F}(N(t)))
\end{aligned}\] fa \[\begin{aligned}
f_i(t)&\equiv f(t_{i+1}-t_i|\mathcal{F}(N(t_i)))\\
&=\Lambda(t_i,t|\mathcal{F}(N(t_i))\exp\left(-t\Lambda(t_i,t|\mathcal{F}(N(t_i))\right)
\end{aligned}\]

    It follows that the log likelihood \(\ell_i:=\log f_i\) of such a
waiting time \(s_i\) is given

\[\begin{aligned}
\ell_i(t_{i+1}-t_i) &=\log \Lambda(s_|\mathcal{F}(N(t_i)) - s_i \Lambda(s_i|\mathcal{F}(N(t_i))\\
&= \log \int_{t_i}^{t_{i+1}} \lambda(s|\mathcal{F}(N(t_i))) ds - {t_{i+1}} \int_{t_i}^{t_{i+1}} \lambda(s|\mathcal{F}(N(t_i))) ds
\end{aligned}
\]

    We may therefore write the likelihood for the data \[
\begin{aligned}
\ell(\mathbf{t}) &= \sum_{1\leq i\leq N(T)}\ell_i(t_{i+1}-t_i)\\
&= \sum_{1\leq i\leq N(T)} \left( \log \int_{t_i}^{t_{i+1}} \lambda(s|\mathcal{F}(N(t_i))) ds - {t_{i+1}} \int_{t_i}^{t_{i+1}} \lambda(s|\mathcal{F}(N(t_i))) ds\right)\\
&= \sum_{1\leq i\leq N(T)} \log \int_{t_i}^{t_{i+1}} \lambda(s|\mathcal{F}(N(t_i))) ds - \sum_{1\leq i\leq N(T)} {t_{i+1}} \int_{t_i}^{t_{i+1}} \lambda(s|\mathcal{F}(N(t_i))) ds\\
&= \sum_{1\leq i\leq N(T)} \log \int_{t_i}^{t_{i+1}} \lambda(s|\mathcal{F}(N(t_i))) ds - \sum_{1\leq i\leq N(T)} {t_{i+1}} \int_{t_i}^{t_{i+1}} \lambda(s|\mathcal{F}(N(t_i))) ds
\end{aligned}
\]

    The Hawkes process \cite{hawkes_point_1971} in particular, is the linear
self-exciting point process; Concretely, given a "background" rate
\(\mu:\mathbb{R}^+\rightarrow\mathbb{R}^+,\) and a non-negative
influence kernel with positive support
\(\phi:\mathbb{R}\rightarrow\mathbb{R}^+\cup\{0\}\), the Hawkes process
has the following expression for its intensity:

\[
\begin{aligned}
\lambda(t|\mathcal{F}(N(t);\mu,\phi) &= \mu(t) + \phi * N\\
&= \mu(t)  + \int_{-\infty}^{\infty}\phi(t-s)dN(s)\\
&= \mu(t) + \int_{-\infty}^{t}\phi(t-s)dN(s)\\
&= \mu(t) + \sum_{t_i\lt t}\phi(t-t_i)
\end{aligned}
\]

For brevity, we write
\(\lambda(\mathcal{F}(N(t));\mu,\phi)=\lambda^*(t)\), but will suppress
explicity dependence on history and parameters..

    The log-likelihood function for a given realisation of the data is can
be found from the usual likelihood for Poisson data of a given intensity
\cite{ozaki_etimating_1979}

\[\begin{aligned}L(N;\mu,\phi) &=-\int_0^T\lambda^*(t)dt + \int_0^T\log \lambda^*(t) dN_t \\
&=-\int_0^T\lambda^*(t)dt + \int_0^T\log \lambda^*(t) dN_t \\
&=-\int_0^T\lambda^*(t)dt + \sum_{t\in\mathbf{t}\cap [0,T]}\log \lambda^*(t)
\end{aligned}\]

Using the method of maximum likelihood we estimate our parameter vector
\(\theta:=(\mu, \phi)\) by choosing the estimates to maximise the
likelihood,
\[\hat{\theta}_\pi(N) = \operatorname{argmax}_{\mu,\phi} L_\pi(N;\theta).\]

    Here we extend the likelihood approach, treating it as penalised
functional regression problem. We augment the log-likeihood function
with a penalty functional \(\pi\):

@@

Similarly, we will approximate \(\phi\) in a different functional basis
\(\{\bar{\psi}\}_{1\leq k \leq K}\), such that
\(\phi(t):= \sum_{k=1}^K \bar{\psi}(t;k)\). This basis choice is likely
arbitrary, but many exist in the literature trading off computational
and statistical convenience, including Laguerre polynomials
\cite{ogata_linear_1982}, exponential functions
\cite{schoenberg_consistent_2005,rambaldi_modeling_2015}, and step
functions \cite{eichler_graphical_2016} As a convenient choice, we
propose a basis of smooth unimodal non-negative functions with a
computationally convenient form, the Maxwell density,
\[\bar{\psi}_j(t;\sigma):=\sqrt{\frac{2}{\pi}} \frac{t^2\exp\left(-\frac{t^2}{2\sigma^2}\right)}{\sigma^3}.\]
so that

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_15_0.pdf}
    \end{center}
    { \hspace*{\fill} \\}

@@

    With basis functions fixed, the problem has become one of selection of
intercepts and basis expansions coefficients, and we may equivalently
write
\(\theta=(\mu, \kappa_1,\kappa_2,\dots,\kappa_K,\omega_1, \omega_2,\dots,\omega_M).\)

\(\ell_1\) type penalties are known for their ability to produce sparse
solutions, so we choose penalties of the form
\[\pi(\theta)=p\left\|\{\kappa_j\}\right\|_1+q\left\|\{\omega_j\}\right\|_1.\]
With the right choice of the parameters \(p\) and \(q\) we hope to
shrink unnecessary parameters to zero, identifying predictively
necessary ones, as in Lasso selection.

    There is no closed-form solution for the optimum of the Hawkes model
likelihood, penalised or not. However, we may use the gradient to find
it efficiently numerically. The optimisation proceeds using a standard
gradient-based black-box based Truncated Newton algorithm; as more
specialised sparse regression algorithms such as Least-Angle Regression,
do not directly apply here.

The gradient is efficient to calculate using standard standard
reverse-mode automatic differentiation \cite{baydin_automatic_2014} for
a wide variety of functional decomposition bases and penalties.

The enthusiast may also calculate the derivatives analytically. For
example, Ozaki \cite{ozaki_maximum_1979} give derivatives for the
parametric log likelihood estimator.

    Let \(\gamma\) be any of the components of the parameter vector
\(\theta\). Since the likelihood depends on this vector only through the
intensity process, we have

\[\begin{aligned}
{ \scriptstyle \frac{\partial}{\partial\gamma} } L_\pi(N;\theta)&=-\int_0^T{ \scriptstyle \frac{\partial}{\partial\gamma} }\lambda^*(t;\theta)dt + \int_0^T { \scriptstyle \frac{\partial}{\partial\gamma} }\log \lambda^*(t;\theta) dN_t -{ \scriptstyle \frac{\partial}{\partial\gamma} }\pi(\theta) \\
&=-\int_0^T{ \scriptstyle \frac{\partial}{\partial\gamma} }\lambda^*(t;\theta)dt + \sum_{0\leq t_i\leq T} \frac{{ \scriptstyle \frac{\partial}{\partial\gamma} }\lambda^*(t_i)}{\lambda^*(t_i)} -{ \scriptstyle \frac{\partial}{\partial\gamma} }\pi(\theta)
\end{aligned}
\]

    Taking \(\gamma=\mu\),

\[{ \scriptstyle \frac{\partial}{\partial\mu} }\lambda^*(t;\mu) = 1\]

I use higher derivatives for \(\mu\) so that I may optimize this
component using a higher order Newton method, since we know that its
optimum is typically shallow \cite{veen_estimation_2008}.

\[\begin{aligned}
  { \scriptstyle \frac{\partial}{\partial\mu} } L_\pi(N;\theta) &=-T + \sum_{0\leq t_i\leq T} \frac{1}{\lambda(t_i;\theta)} \\
  { \scriptstyle \frac{\partial^2}{\partial\mu^2} } L_\pi(N;\theta) &= \sum_{0\leq t_i\leq T} \frac{-1}{\lambda^2(t_i;\theta)}\\
  { \scriptstyle \frac{\partial^3}{\partial\mu^3} } L_\pi(N;\theta) &= \sum_{0\leq t_i\leq T} \frac{2}{\lambda^3(t_i;\theta)}
  \end{aligned}\]

    Now I handle \(\gamma=\kappa_j\),

\[{ \scriptstyle \frac{\partial}{\partial\kappa_j} }\lambda^*(t;\theta) =  \sum_{t_i\lt t}\phi_j(t-t_i)\]

so that

\[\begin{aligned}
  { \scriptstyle \frac{\partial}{\partial\kappa_i} } L_\pi(N; \theta)&=-\int_0^T \sum_{t_i\lt t}\phi_i(t-t_i) dt + \sum_{0\leq t_i\leq T} \frac{\sum_{t_k\lt t_i}\phi_j(t_i-t_k)}{\lambda_\theta(t_i)} -\frac{\partial}{\partial\kappa_j}\pi(\theta)
\end{aligned}\]

    Finally, I handle the \(\omega_j\) values.

Taking \(\theta=\omega_j\), and defining
\(\Delta\tau_j:= {\tau_{j-1}}-{\tau_j}\) we find

\[
\begin{align*}
{ \scriptstyle \frac{\partial}{\partial\omega_j} }\pi\|\boldsymbol \omega_\theta(t)\|_1 &= { \scriptstyle \frac{\partial}{\partial\omega_j} }\pi \int_{\tau_{j-1}}^{\tau_j} \left|\omega_j\right|dt\\
&=\pi\Delta\tau_j\operatorname{sgn} \omega_j\\
{ \scriptstyle \frac{\partial}{\partial\omega_j} }\lambda_\theta(t|\mathcal{F}_t) &=  \Delta\tau_j\mathbb{I}_{[\tau_{j-1},\tau_j)}(t)\\
{ \scriptstyle \frac{\partial}{\partial\omega_j} } L_\pi(\mathbf t;\theta) &=-\Delta\tau_j-\pi\Delta\tau_j\operatorname{sgn} \omega_j + \sum_{\tau_{j-1}\leq t_i\leq \tau_j} \frac{1}{\lambda_\theta(t_i)}
\end{align*}
\]

Higher partial derivatives are also analogous to the partial derivatives
with respect to \(\mu\), although of course the penalty introduces a
discontinuity at \(\omega_j=0\). This last formula is the key to the
gradient ascent algorithm.

Note that the \(\omega_j\) values are mutually orthogonal, in the sense
that \({ \scriptstyle \frac{\partial^2}{\partial\omega_i\omega_j}} =0\)
if \(i\ne j\). I can treat these components, more or less, as separate
univariate components when optimizing, and the Hessian will be sparse
off the diagonal.

By construction,
\({ \scriptstyle \frac{\partial}{\partial\mu} }\pi(\gamma) = { \scriptstyle \frac{\partial}{\partial\eta} }\pi(\gamma) = { \scriptstyle \frac{\partial}{\partial\kappa} }\pi(\gamma) = 0\).

TODO: I've changed \(\omega\) penalisation; this needs fixing.

@@

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_31_0.pdf}
    \end{center}
    { \hspace*{\fill} \\}

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_31_1.pdf}
    \end{center}
    { \hspace*{\fill} \\}

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_31_2.pdf}
    \end{center}
    { \hspace*{\fill} \\}

@@

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_33_0.png}
    \end{center}
    { \hspace*{\fill} \\}

@@

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_35_0.png}
    \end{center}
    { \hspace*{\fill} \\}
